Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-07-08 submission/manuscript.Rmd
discussion formatting

A single genome holds the instructions to generate the myriad of cell types found within the adult organism. This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001]. Although many chromatin regulators were initially identified for their roles in shaping cellular and tissue identity[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a], recent advancements in next generation sequencing unexpectedly revealed their dysregulation underlies many neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018]. Many studies have suggested this connection between neurodevelopment and chromatin regulators is due to their regulation of brain-specific genes, such as orchestrating the transcriptional programs for synaptic maturation[@zhouBrainspecificPhosphorylationMeCP22006] and the transition between neuronal and glial differentiation in neural precursor cells[@hirabayashiPolycombLimitsNeurogenic2009]. However, loss of some chromatin regulators can also result in the ectopic transcription of tissue-specific genes outside of their target environment[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017]. Very few studies have investigated the misexpression of tissue-specific genes in chromatin-linked neurodevelopmental disorders[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is currently unclear if this partial loss of brain identity contributes to neurodevelopmental impairments.

@@

# Discussion
- The demarcation of the germ vs soma is a key feature of multicellularity
- Other H3K4me regulators, anything known about tissue-biased gene expression?
- This suggests the germline developmental program is occuring ectopically as _Kdm5c_-KOs progresses through somatic tissue development
- tissue-biased gene expression:
	- However unlike the gonadal-biased DEGs, many liver and muscle-biased DEGs have a known involvement in brain function. For example, the liver-biased DEG is _Apolipoprotein C-I (Apoc1)_, is important for lipoprotein metabolism but has also been shown to influence learning and memory (Figure 1E).
- Otx2 is properly expressed in _Kdm5c_-KO EpiLCs and is known to prevent pgc identity https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6485399/

@@

* Papers to read/reference:
	* Reconstitution of the Mouse Germ Cell Specification Pathway in Culture by Pluripotent Stem Cells: https://www.cell.com/fulltext/S0092-8674(11)00771-9
	* two cell gene list used by Suzuki et al Max paper is based on 2 cell sequencing: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3395470/

>>>>

A single genome holds the instructions to generate all the myriad of cell types found within the adult organism. This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001]. Although many chromatin regulators were initially identified for their roles in shaping cellular and tissue identity[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a], recent advancements in next generation sequencing unexpectedly revealed their dysregulation underlies many neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018]. Many studies have suggested this connection between neurodevelopment and chromatin regulators is due to their regulation of brain-specific genes, such as orchestrating the transcriptional programs for synaptic maturation[@zhouBrainspecificPhosphorylationMeCP22006] and the transition between neuronal and glial differentiation in neural precursor cells[@hirabayashiPolycombLimitsNeurogenic2009]. However, loss of some chromatin regulators can also result in the ectopic transcription of tissue-specific genes outside of their target environment[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017]. Very few studies have investigated the misexpression of tissue-specific genes in chromatin-linked neurodevelopmental disorders[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is currently unclear if this partial loss of brain identity contributes to neurodevelopmental impairments.

@@

# Discussion
Random thoughts
- The demarcation of the germ vs soma is a key feature of multicellularity
- Anything known about tissue-biased gene expression in other H3K4me regulators?
- Our data suggests the germline developmental program is occuring ectopically as _Kdm5c_-KOs progresses through somatic tissue development
- tissue-biased gene expression:
	- However unlike the gonadal-biased DEGs, many liver and muscle-biased DEGs have a known involvement in brain function. For example, the liver-biased DEG is _Apolipoprotein C-I (Apoc1)_, is important for lipoprotein metabolism but has also been shown to influence learning and memory.
- Otx2 is expressed in EpiLCs and is known to repress PGC identity. 
	- It's properly expressed in _Kdm5c_-KO EpiLCs, further supporting they aren't just becoming PGCs https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6485399/

@@

- Papers to read/reference:
	- Reconstitution of the Mouse Germ Cell Specification Pathway in Culture by Pluripotent Stem Cells: https://www.cell.com/fulltext/S0092-8674(11)00771-9
	- two cell gene list used by Suzuki et al Max paper is based on 2 cell sequencing: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3395470/
